MCC Models

### 2019-07-23 papers/full_paper/SI_appendix.Rmd
continue fleshing our numerical example

The situation we model is defined by three sets: referents $r$, utterances $u$, and lexica $\mathcal{L}$. 
Referents are defined by two features: the type of object they are (visually discernible given their shape and color) and their location (one of two tables).
There are two types of objects $t_1$, $t_2$ and two locations $l_1$ and $l_2$.
The utterances available to a speaker are (action, label) pairs, where the action involves turning (or pointing) to a particular location (a table) and labels are novel words ($w_1$, $w_2$), which are assumed to refer to the type of the object. 
The action of turning or pointing reduces the set of referents to those that are on the table targeted by the turn (point); interpreting a label reduces that set further to those which are of the type referred to by that label.
The interpretation of a label requires a lexicon $\mathcal{L}$, which provides the label--type mappings.
There are two lexica: one which maps label 1 $w_1$ to type 1 $t_1$ and label 2 $w_2$ to type 2 $t_2$, and one which has the inverse mapping.

@@

Our word-learning model is a model of pragmatic reasoning couched in the Rational Speech Act modeling framework. The model describes the following process: A pragmatic listener ($L_1$) jointly infers a referent (what object is being picked out by the utterance) and a lexicon (label--type mappings) by reasoning about a pragmatic speaker ($S_1$) who produces utterances to convey information to a literal listener ($L_0$), who in turn interprets utterances according to their literal meaning. According to Bayes rule, this inference is given by:

@@

The right-hand side of this equation has three components: the prior distribution over referents $P(r)$, the prior distribution over lexica $P(\mathcal{L})$, and the likelihood that a speaker would produce the an utterance $u$ given a referent $r$ ($P_{S1}(u|r)$).
In the situation we model, the prior on referents $P(r)$ is a categorical distribution over three objects in a scene, which we posit could be non-uniform due to what is in common ground (see below for common ground manipulations). 
Because the labels produced by the speaker are all novel words, the listener has no substantive knowledge about the lexica (label--type mappings) and the prior over the two lexica (described above) is uniform.

The listener updates their beliefs about both the referent and the lexicon by reasoning about the speaker, assumed to produce utterances to convey the referent to the listener by being a soft-max rational agent (with degree of rationality $\alpha$) with a utility function defined in terms of the informativity of an utterance for a referent:

@@

The informativity of an utterance for a referent is the (log) probability that a naive listener $L_0$ would select that referent given that utterance.

@@

where $P_{L_0}(r, \mathcal{L}|u)  \propto \mathcal{L}_{lit} P(r) P(\mathcal{L})$.
$P(r)$ again denotes the prior probability of a referent, $\mathcal{L}_{lit}$ encodes the literal meaning of an utterance given a particular lexicon, returning 1 if the utterance is literally true and 0 if the utterance is literally false. As mentioned above, because $L_1$ does not know the lexicon, the semantics of the words contained in the utterance (i.e., the labels) offer no information about the referent. The non-linguistic aspect of the utterance (the turning or pointing), however, do. As described in above, the semantics of turning to one of the tables is roughly equivalent to saying "It's an object on that table". That is, $\mathcal{L}_{lit}$ returns 1 for a referent that is on the table the agent turned to and 0 for referents on the other table.

@@

In the following, we work through a numerical example for how model predictions were generated for the pragmatics model. The prediction will correspond to the parameter free model described above (excluding the noise parameter). The values of the parameters are taken from the preference - different speaker - congruent condition. Fig. 1 shows a screenshot from the adult experiment, which the model was designed to capture. In this context, there are three potential referents (two red-ish: $r_1^r$, $r_2^r$; one yellow-ish, $r_3^y$,), of two types (red-ish and yellow-ish) on two tables. 
In principle, the speaker, the frog in this case, can perform one of four utterances, turning either to the left or right table and producing either label ("dax" or "wug"):  $u_1 = (\text{left}, dax)$, $u_2 = (\text{left}, wug)$, $u_3 = (\text{right}, dax)$, $u_4 = (\text{right}, wug)$.
We work out the example where the speaker produces utterance $u_3$, turning to the right table (two objects) and saying the label "dax".

@@

MH: SHOULD MAKE THIS A DISTRIBUTION (sums to 1).

The prior distribution over referents $P(r)$ (left to right in Fig.1), based on Experiment 2A, for this condition was [`r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`]. That is, the yellow object has a higher a priori probability compared to red objects, because the speaker previously displayed a preference for a yellow object.

The literal listener's posterior distribution over referents and lexica a support of size six, owing to the unique combinations of referents and lexica. 
$$
\begin{aligned}
& P_{L_0}(r_1^r, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_1^r, u_3) P(r_1^r) P(L_1) = 0 \times 0.36 \times 0.5\\
& P_{L_0}(r_2^r, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_2^r, u_3) P(r_2^r) P(L_1) = 1 \times 0.36 \times 0.5\\
& P_{L_0}(r_3^y, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_3^y, u_3) P(r_3^y) P(L_1) = 0 \times 0.64 \times 0.5\\
& P_{L_0}(r_1^r, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_1^r, u_3) P(r_1^r) P(L_2) = 0 \times 0.36 \times 0.5\\
& P_{L_0}(r_2^r, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_2^r, u_3) P(r_2^r) P(L_2) = 0 \times 0.36 \times 0.5\\
& P_{L_0}(r_3^y, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_3^y, u_3) P(r_3^y) P(L_2) = 1 \times 0.64 \times 0.5\\
\end{aligned}
$$

because the speaker is pointing to the table with $r_2$ and $r_3$, and under $\mathcal{L_1}$, "dax" means the red object and under $\mathcal{L_2}$, "dax" means the yellow object.
After normalization we have:

$$
\begin{aligned}
& P_{L_0}(r_1^r, \mathcal{L}_1 | {u_3}) = 0 \\
& P_{L_0}(r_2^r, \mathcal{L}_1 | {u_3}) = 0.36 \\
& P_{L_0}(r_3^y, \mathcal{L}_1 | {u_3})   = 0 \\
& P_{L_0}(r_1^r, \mathcal{L}_2 | {u_3})   = 0 \\
& P_{L_0}(r_2^r, \mathcal{L}_2 | {u_3})   = 0 \\
& P_{L_0}(r_3^y, \mathcal{L}_2 | {u_3}) = 0.64\\
\end{aligned}
$$

@@

$$
\begin{aligned}
&P_{L_0}(r_1|u_3) =  P_{L_0}(r_1, \mathcal{L_1}|u_3) + P_{L_0}(r_1, \mathcal{L_2}|u_3) = 0+0=0 \\
&P_{L_0}(r_2|u_3) =  P_{L_0}(r_2, \mathcal{L_1}|u_3) + P_{L_0}(r_2, \mathcal{L_2}|u_3) = 0.36+0=0.36 \\
&P_{L_0}(r_3|u_3) =  P_{L_0}(r_3, \mathcal{L_1}|u_3) + P_{L_0}(r_3, \mathcal{L_2}|u_3) = 0+64=0.64 \\
\end{aligned}
$$

@@

$$
\begin{aligned}
&Informativity(u_3; r_1) = log(0) = -\infty \\
&Informativity(u_3; r_2) = log(0.36) = -1.03 \\
&Informativity(u_3; r_3) = log(0.64) = -0.44
\end{aligned}
$$

@@

SEEMS LIKE WE NEED TO NORMALIZE ACROSS ALL UTTERANCES, SO WE WILL NEED TO REDO the ABOVE CALCULATIONS FOR EACH OF THE 6 UTTERANCES.

>>>>

The situation we model is defined by three sets: referents $r$, utterances $u$, and lexica $\mathcal{L}$. 
Referents are defined by two features: the type of object they are (visually discernible given their shape and color) and their location (one of two tables).
There are two types of objects $t_1$, $t_2$ and two locations $l_1$ and $l_2$.
The utterances available to a speaker are (action, label) pairs, where the action involves turning (pointing) to a particular location (a table) and labels are novel words ($w_1$, $w_2$), which are assumed to refer to the type of the object. 
The action of turning (pointing) reduces the set of referents to those that are on the table targeted by the turn (point); interpreting a label reduces that set further to those which are of the type referred to by that label.
The interpretation of a label requires a lexicon $\mathcal{L}$, which provides the label--type mappings.
There are two lexica: one which maps label 1 $w_1$ to type 1 $t_1$ and label 2 $w_2$ to type 2 $t_2$, and one which has the inverse mapping.

@@

Our word-learning model is a model of pragmatic reasoning couched in the Rational Speech Act modeling framework. The model describes the following process: A pragmatic listener ($L_1$) jointly infers a referent (what object is being picked out by the utterance) and a lexicon (label--type mappings) by reasoning about a pragmatic speaker ($S_1$) who produces utterances to convey information to a literal listener ($L_0$), who in turn interprets utterances according to their literal meaning. According to Bayes rule, the pragmatic listener's inference is given by:

@@

The right-hand side of this equation has three terms: the prior distribution over referents $P(r)$, the prior distribution over lexica $P(\mathcal{L})$, and the likelihood that a speaker would produce an utterance $u$ given a referent $r$ ($P_{S1}(u|r)$).
In the situation we model, the prior on referents $P(r)$ is a categorical distribution over three objects in a scene, which we posit could be non-uniform due to what is in common ground (see below for common ground manipulations). 
Because the labels produced by the speaker are all novel words, the listener has no substantive knowledge about the lexica (label--type mappings) and thus the prior over the two lexica (described above) is uniform.

The pragmatic listener updates their beliefs about both the referent and the lexicon by reasoning about the speaker, assumed to produce utterances to convey the referent to the listener by being a soft-max rational agent (with degree of rationality $\alpha$) with a utility function defined in terms of the informativity of an utterance for a referent:

@@

As spelled out in @GoodmanStuhlmuller2013, the informativity of an utterance for a referent is the (log) probability that a naive listener $L_0$ would select that referent given that utterance.

@@

where $P_{L_0}(r, \mathcal{L}|u)  \propto \mathcal{L}_{lit} P(r) P(\mathcal{L})$.
$P(r)$ again denotes the prior probability of a referent, $\mathcal{L}_{lit}$ encodes the literal meaning of an utterance given a particular lexicon, returning 1 if the utterance is literally true and 0 if the utterance is literally false. As mentioned above, because $L_1$ does not know the lexicon, the semantics of the words contained in the utterance (i.e., the labels) offer no information about the referent. The non-linguistic aspect of the utterance (the turning or pointing), however, do. As described above, the semantics of turning to one of the tables is roughly equivalent to saying "It's an object on that table". That is, $\mathcal{L}_{lit}$ returns 1 for a referent that is on the table the agent turned to and 0 for referents on the other table.

@@

In this section, we work through a toy numerical example for how model predictions were generated for the pragmatics model. The prediction will correspond to the parameter free model described above (excluding the noise parameter). The values of the parameters are taken from the preference - different speaker - congruent condition (see below). Fig. 1 shows a screenshot from the adult experiment, which the model was designed to capture. In this context, there are three potential referents (two red-ish: $r_1^r$, $r_2^r$; one yellow-ish, $r_3^y$,), of two types (red-ish and yellow-ish) on two tables. 
In principle, the speaker, the frog in this case, can produce one of four utterances, turning either to the left or right table and saying either label ("dax" or "wug"):  $u_1 = (\text{left}, dax)$, $u_2 = (\text{left}, wug)$, $u_3 = (\text{right}, dax)$, $u_4 = (\text{right}, wug)$.
We work out the example where the speaker produces utterance $u_3$, turning to the right table (two objects) and saying the label "dax" (though since the labels have no *a priori* meanings, the computation would be the same for utterance $u_4$). <-- CHECK ME.

@@

We construct the prior distribution over referents $P(r)$ based on the results of Experiment 2A, in which the speaker displayed a preference for a yellow object (see Model Parameters section below for a detailed description of how this distribution was constructed). 
The prior distribution over referents $P(r)$ (left to right in Fig.1) for this condition was [0.26, 0.26, 0.48].
<!-- [`r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(1-mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`; `r round(mean(adult_ex2_preference_data%>%filter(condition == "different_speaker")%>%pull(correct)),2)`].  -->

The literal listener's posterior distribution over referents and lexica has support of size six, owing to the unique combinations of referents and lexica.
$$
\begin{aligned}
& P_{L_0}(r_1^r, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_1^r, u_3) P(r_1^r) P(L_1) = 0 \times 0.26 \times 0.5\\
& P_{L_0}(r_2^r, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_2^r, u_3) P(r_2^r) P(L_1) = 1 \times 0.26 \times 0.5\\
& P_{L_0}(r_3^y, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_3^y, u_3) P(r_3^y) P(L_1) = 0 \times 0.48 \times 0.5\\
& P_{L_0}(r_1^r, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_1^r, u_3) P(r_1^r) P(L_2) = 0 \times 0.26 \times 0.5\\
& P_{L_0}(r_2^r, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_2^r, u_3) P(r_2^r) P(L_2) = 0 \times 0.26 \times 0.5\\
& P_{L_0}(r_3^y, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_3^y, u_3) P(r_3^y) P(L_2) = 1 \times 0.48 \times 0.5\\
\end{aligned}
$$

because the speaker is pointing to the table with $r_2$ and $r_3$, and under $\mathcal{L}_1$, "dax" means the red object and under $\mathcal{L}_2$, "dax" means the yellow object.
After normalization we have:

$$
\begin{aligned}
& P_{L_0}(r_1^r, \mathcal{L}_1 | {u_3}) = 0 \\
& P_{L_0}(r_2^r, \mathcal{L}_1 | {u_3}) = 0.35 \\
& P_{L_0}(r_3^y, \mathcal{L}_1 | {u_3})   = 0 \\
& P_{L_0}(r_1^r, \mathcal{L}_2 | {u_3})   = 0 \\
& P_{L_0}(r_2^r, \mathcal{L}_2 | {u_3})   = 0 \\
& P_{L_0}(r_3^y, \mathcal{L}_2 | {u_3}) = 0.65\\
\end{aligned}
$$

@@

$$
\begin{aligned}
&P_{L_0}(r_1|u_3) =  P_{L_0}(r_1, \mathcal{L}_1|u_3) + P_{L_0}(r_1, \mathcal{L}_2|u_3) = 0+0=0 \\
&P_{L_0}(r_2|u_3) =  P_{L_0}(r_2, \mathcal{L}_1|u_3) + P_{L_0}(r_2, \mathcal{L}_2|u_3) = 0.35+0=0.35 \\
&P_{L_0}(r_3|u_3) =  P_{L_0}(r_3, \mathcal{L}_1|u_3) + P_{L_0}(r_3, \mathcal{L}_2|u_3) = 0+0.65=0.65 \\
\end{aligned}
$$

@@

$$
\begin{aligned}
&Informativity(u_3; r_1) = log(0) = -\infty \\
&Informativity(u_3; r_2) = log(0.35) = -1.04 \\
&Informativity(u_3; r_3) = log(0.65) = -0.43
\end{aligned}
$$

@@

SEEMS LIKE WE NEED TO NORMALIZE ACROSS ALL UTTERANCES, SO WE WILL NEED TO REDO THE ABOVE CALCULATIONS FOR EACH OF THE 6 UTTERANCES.
